---
title: "DATA VISUALIZATION"
---

## FINANCIAL PLANNING VISUALIZATION

<div class="brutal-box-green">
**PROJECT CONTEXT**

This visualization was created for my **Fintech Finance Course** at SIT, demonstrating the long-term impact of compound interest on Singapore's Central Provident Fund (CPF) savings.
</div>

### THE VISUALIZATION
```{r}
#| echo: false
#| warning: false
#| message: false

library(ggplot2)

# Age range
ages <- 25:65

# Parameters
starting_salary <- 3500
increment_rate <- 0.03
oa_rate <- 0.025  # Ordinary Account interest
sa_rate <- 0.04   # Special Account interest
ma_rate <- 0.04   # MediSave Account interest

# Initialize balances
oa_balance <- numeric(length(ages))
sa_balance <- numeric(length(ages))
ma_balance <- numeric(length(ages))

oa_balance[1] <- 0
sa_balance[1] <- 0
ma_balance[1] <- 0

# Calculate year-by-year balances
for (i in 2:length(ages)) {
  year <- i - 1
  salary <- starting_salary * (1 + increment_rate)^year

  # Monthly contributions (simplified CPF rates)
  monthly_oa <- salary * 0.23
  monthly_sa <- salary * 0.16
  monthly_ma <- salary * 0.08

  # Annual contributions
  annual_oa <- monthly_oa * 12
  annual_sa <- monthly_sa * 12
  annual_ma <- monthly_ma * 12

  # Apply interest and add contributions
  oa_balance[i] <- oa_balance[i-1] * (1 + oa_rate) + annual_oa
  sa_balance[i] <- sa_balance[i-1] * (1 + sa_rate) + annual_sa
  ma_balance[i] <- ma_balance[i-1] * (1 + ma_rate) + annual_ma
}

# Create data frame
df <- data.frame(
  Age = rep(ages, 3),
  Balance = c(oa_balance, sa_balance, ma_balance),
  Account = rep(c("Ordinary Account", "Special Account", "MediSave Account"),
                each = length(ages))
)

# Create the visualization
ggplot(df, aes(x = Age, y = Balance, color = Account, group = Account)) +
  geom_line(size = 1.2) +
  scale_color_manual(values = c("Ordinary Account" = "#3498db",
                                  "Special Account" = "#e74c3c",
                                  "MediSave Account" = "#2ecc71")) +
  scale_y_continuous(labels = scales::dollar_format(prefix = "$", big.mark = ",")) +
  labs(
    title = "CPF Account Balance Projections Over Time",
    x = "Age (years)",
    y = "Account Balance (SGD)",
    color = "Account Type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 15, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 12, face = "bold"),
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

<hr>

## WHAT WAS VISUALIZED

<div class="brutal-grid-3">
<div class="brutal-box">
**X-AXIS**

Age in years (25 to 65) - continuous numeric variable representing career progression
</div>

<div class="brutal-box">
**Y-AXIS**

Account balance in SGD - monetary values showing accumulated savings over time
</div>

<div class="brutal-box">
**THREE SERIES**

Ordinary, Special, MediSave Accounts - each representing different CPF account types with distinct interest rates
</div>
</div>

### SAMPLE NUMERICAL DATA
```{r}
#| echo: false

# Create sample table with proper formatting
sample_ages <- c(25, 30, 35, 40, 45, 50, 55, 60, 65)
sample_indices <- sample_ages - 24

sample_data <- data.frame(
  Age = sample_ages,
  Ordinary = sprintf("$%s", format(round(oa_balance[sample_indices]), big.mark = ",", trim = TRUE)),
  Special = sprintf("$%s", format(round(sa_balance[sample_indices]), big.mark = ",", trim = TRUE)),
  MediSave = sprintf("$%s", format(round(ma_balance[sample_indices]), big.mark = ",", trim = TRUE))
)

knitr::kable(sample_data,
             col.names = c("Age", "Ordinary Account", "Special Account", "MediSave Account"),
             align = c("r", "r", "r", "r"))
```

<hr>

## DATA SOURCE AND ASSUMPTIONS

<div class="brutal-grid-2">
<div class="brutal-box">
**OFFICIAL RATES**

— CPF contribution rates from CPF Board (2024)

— Interest rates: OA (2.5% p.a.), SA (4% p.a.), MA (4% p.a.)

— Based on actual Singapore government policies
</div>

<div class="brutal-box">
**CALCULATION ASSUMPTIONS**

— Starting monthly salary: SGD $3,500

— Annual salary increment: 3%

— Contribution percentages: OA (23%), SA (16%), MA (8%)

— Time horizon: 40 years (age 25-65)

— No voluntary contributions or withdrawals
</div>
</div>

<hr>

## RATIONALE FOR VISUALIZATION METHOD

<div class="brutal-box-green">
**WHY LINE CHART?**

I chose a **line chart** because it's the optimal choice for displaying time-series financial data with multiple comparisons.
</div>

<div class="brutal-grid-2">
<div class="brutal-box">
**ADVANTAGES OF LINE CHART**

**Time-series excellence:** Shows how numeric values change over continuous time periods

**Trend visibility:** Upward curves clearly show exponential growth - steeper slopes in later years are immediately visible

**Multiple comparisons:** Three distinct colors enable easy comparison of different account growth rates

**Financial convention:** Standard for investment growth visualization, making interpretation intuitive
</div>

<div class="brutal-box">
**WHY NOT ALTERNATIVES**

**Bar chart:** Would create visual clutter with 40+ bars per account type

**Stacked area chart:** Makes it difficult to read individual account values accurately

**Scatter plot:** Would suggest discrete data points rather than continuous progression

**Pie chart:** Cannot show changes over time
</div>
</div>

<hr>

## KEY INSIGHT DISCOVERED

<div class="brutal-box-yellow">
The visualization revealed something surprising: **the last 10 years of savings (age 55-65) contribute almost as much absolute value as the first 20 years (age 25-45)**.

**Why?** Compound interest. By year 30, you're earning interest on decades of contributions AND their accumulated interest. The raw numbers showed this, but the **upward curving lines** made it intuitively clear.
</div>

<hr>

## KEY INSIGHTS FROM THE DATA

<div class="brutal-grid-3">
<div class="brutal-box-green">
**COMPOUND INTEREST POWER**

**Age 30** (5 years): Total ≈ $115,000

**Age 50** (25 years): Total ≈ $800,000

**INSIGHT:** The middle 20 years contribute **7x more** than the first 5 years
</div>

<div class="brutal-box-pink">
**ACCOUNT TYPE DIFFERENCES**

**Special Account:** 4% interest

**Ordinary Account:** 2.5% interest

**TAKEAWAY:** This 1.5% difference creates over **$80,000 additional value** over 40 years
</div>

<div class="brutal-box-yellow">
**EARLY VS LATE CONTRIBUTIONS**

**$1,000 at age 25:** grows to ~$2,685 (2.69x)

**$1,000 at age 45:** grows to ~$1,480 (1.48x)

**CONCLUSION:** Starting early provides **82% more growth** per dollar invested
</div>
</div>

<hr>

## EDUCATIONAL VALUE

<div class="brutal-grid-2">
<div class="brutal-box">
**DATA VISUALIZATION PRINCIPLES**

Learned how to choose appropriate chart types for quantitative time-series data and apply effective visual encoding.
</div>

<div class="brutal-box">
**COMPOUND INTEREST MATHEMATICS**

Understood the exponential nature of investment growth through concrete numerical examples.
</div>

<div class="brutal-box">
**EFFECTIVE COMMUNICATION**

Mastered transforming raw numerical data into clear, actionable insights that non-technical audiences can understand.
</div>

<div class="brutal-box">
**FINANCIAL LITERACY**

Gained deep understanding of Singapore's CPF system through hands-on quantitative modeling and analysis.
</div>
</div>

<hr>

## CONCLUSION

**Early contributions have disproportionate impact** - visible in curve accumulation patterns

**Higher interest accounts grow significantly faster** - visible in progressive line separation

**Growth accelerates exponentially over time** - visible in steepening curves toward retirement age

The key achievement: converting hundreds of individual numeric calculations into a single comprehensible story about the mathematical power of compound interest and the importance of early financial planning.
